Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-09 docs/Manuscript/Manuscript.Rmd
transcribe revisions to discussion + format performance table + update figures

Here we describe our review process, including deviations from plans and the reasons for them.
More details are included in Section \label{sec:appendix-selection}.

@@

**Correct "n = PubMed" in PRISMA diagram.**
We include a PRISMA checklist as Supplemental Table 1 and a PRISMA abstract checklist as Supplemental Table 2.
Because we focus on methodologies rather than conceptual approaches or evidence, we deviate in some ways from PRISMA guidelines. In particular, those items of the checklist involving bias assessment and quantitative synthesis are intended for meta-analyses so did not apply to this study.
Because we are not aware of any standard procedures for conducting reviews and syntheses of methodology, no protocol was prepared for this study.

@@

**Update numbers below, perhaps dynamically.**

The 25 included studies were analyzed based on 20 characteristics (see Section \ref{sec:data}), and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, and _Hindawi - Journal of Healthcare Engineering_ published 3.

@@

Some PDFs were obtained using Zotero from a university workstation, and the remaining were obtained through university library services. The review process is detailed in Section \label{sec:full-text}.

@@

```{r tab:performance, echo=FALSE, message=FALSE, warning=FALSE}
tab_eval_comp |> 
  pivot_wider(
    id_cols = c(zotero_key, date_of_publication, measure, task),
    names_from = role, values_from = results
  ) |> 
  arrange(date_of_publication) |> 
  mutate(len = map2_int(proposal, comparator, ~ max(length(.x), length(.y)))) |> 
  mutate(across(c(proposal, comparator), ~ map2(., len, \(x, y) x[seq(y)]))) |> 
  unnest(c(proposal, comparator)) |> 
  separate(proposal, c("proposal", " "), sep = ": ") |> 
  separate(comparator, c("comparator", "  "), sep = ": ") |> 
  # use abbreviations throughout
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Ll]ogistic [Rr]egression", "LR")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ ifelse(. == "Logistic", "LR", .)
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Rr]andom [Ff]orest", "RF")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Kk][\\- ][Nn]earest [Nn]eighbor(s){0,1}", "KNN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "([Aa]rtificial ){0,1}[Nn]eural [Nn]et(work){0,1}", "NN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Oo]ne-class", "One-class")
  )) |> 
  mutate(across(
    c(proposal, ` `, comparator, `  `),
    ~ ifelse(is.na(.), "", .)
  )) |> 
  transmute(
    Citation = str_c("[@", zotero_key, "]"),
    Measure = measure, Problem = task,
    Proposals = proposal, ` `,
    # insert blank column to separate right- and left-justified entries
    # `   ` = "",
    Comparators = comparator, `  `
  ) |> 
  mutate(across(
    c(Citation, Measure, Problem),
    ~ ifelse(is.na(lag(.)) | . != lag(.), ., "")
  )) |>
  # prevent numeric list columns from dominating the horizontal space
  # mutate(across(c(Proposals, Comparators), ~ str_c("\\hspace{3em}", .))) |> 
  mutate(across(c(` `, `  `), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "llllrlr",
    caption = paste0(
      "\\label{tab:performance}",
      "Evaluations of proposed methods and comparisons to alternative methods.",
      " (See original studies for full names and descriptions.)"
    )
  )
```

>>>>

Here we describe our review process, including deviations from plans and the reasons for them.
More details are included in Section \ref{sec:appendix-selection}.

@@

We include a PRISMA checklist as Supplemental Table 1 and a PRISMA abstract checklist as Supplemental Table 2.
Because we focus on methodologies rather than conceptual approaches or evidence, we deviate in some ways from PRISMA guidelines. In particular, those items of the checklist involving bias assessment and quantitative synthesis are intended for meta-analyses so did not apply to this study.
Because we are not aware of any standard procedures for conducting reviews and syntheses of methodology, no protocol was prepared for this study.

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on 20 characteristics (see Section \ref{sec:data}), and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3, and _PLOS One_ published 2.

@@

Some PDFs were obtained using Zotero from a university workstation, and the remaining were obtained through university library services. The review process is detailed in Section \ref{sec:full-text}.

@@

```{r tab:performance, echo=FALSE, message=FALSE, warning=FALSE}
tab_eval_comp |> 
  pivot_wider(
    id_cols = c(zotero_key, date_of_publication, measure, task),
    names_from = role, values_from = results
  ) |> 
  arrange(date_of_publication) |> 
  mutate(len = map2_int(proposal, comparator, ~ max(length(.x), length(.y)))) |> 
  mutate(across(c(proposal, comparator), ~ map2(., len, \(x, y) x[seq(y)]))) |> 
  unnest(c(proposal, comparator)) |> 
  separate(proposal, c("proposal", " "), sep = ": ") |> 
  separate(comparator, c("comparator", "  "), sep = ": ") |> 
  # use abbreviations throughout
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Ll]ogistic [Rr]egression", "LR")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ ifelse(. == "Logistic", "LR", .)
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Rr]andom [Ff]orest", "RF")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Kk][\\- ][Nn]earest [Nn]eighbor(s){0,1}", "KNN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "([Aa]rtificial ){0,1}[Nn]eural [Nn]et(work){0,1}", "NN")
  )) |> 
  mutate(across(
    c(proposal, comparator),
    ~ str_replace(., "[Oo]ne-class", "One-class")
  )) |> 
  mutate(across(
    c(proposal, ` `, comparator, `  `),
    ~ ifelse(is.na(.), "", .)
  )) |> 
  transmute(
    Citation = str_c("[@", zotero_key, "]"),
    Measure = measure, Problem = task,
    Proposals = proposal, ` `,
    # insert blank column to separate right- and left-justified entries
    # `   ` = "",
    Comparators = comparator, `  `
  ) |> 
  mutate(Measure = ifelse(
    is.na(lag(Measure)) | Measure != lag(Measure) | Citation != lag(Citation),
    Measure, ""
  )) |> 
  mutate(across(
    c(Citation, Problem),
    ~ ifelse(is.na(lag(.)) | . != lag(.), ., "")
  )) |>
  # prevent numeric list columns from dominating the horizontal space
  # mutate(across(c(Proposals, Comparators), ~ str_c("\\hspace{3em}", .))) |> 
  mutate(across(c(` `, `  `), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "llllrlr",
    caption = paste0(
      "\\label{tab:performance}",
      "Evaluations of proposed methods and comparisons to alternative methods.",
      " (See original studies for full names and descriptions.)"
    )
  )
```

### 2023-09-09 docs/Manuscript/Manuscript.tex
transcribe revisions to discussion + format performance table + update figures

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} that appears to have been introduced
independently several times. The approach fits ML models, including
regularized families of generalized regression models (GLMs), to cohorts
retrieved via similarity matching. Whereas CBR and NN prediction use
patient similarity to obtain an ideally homogeneous subpopulation
matched to the index case, in localized modeling one seeks a relevant
yet heterogeneous subpopulation, each member of which may deviate from
the index case. A prediction or inference is then obtained from a model
fitted to this \emph{similarity cohort}. The use of similarity cohorts
distinguishes localized modeling from classical GLMs that still dominate
analysis in clinical trials and comparative effectiveness research, as
well as from many cutting-edge ML models that are fitted to whole
populations, while the cohorts' heterogeneity and its utility to the
model distinguish the approach from CBR. In principle, this allows
interpretable models like GLMs to identify and measure the effects of
determining factors that are specific to the index case. In the reviewed
studies, localized modeling shows some promise toward simultaneously
achieving the performance improvements of ML and the interpretability of
classical models.

@@

\textbf{Correct ``n = PubMed'' in PRISMA diagram.} We include a PRISMA
checklist as Supplemental Table 1 and a PRISMA abstract checklist as
Supplemental Table 2. Because we focus on methodologies rather than
conceptual approaches or evidence, we deviate in some ways from PRISMA
guidelines. In particular, those items of the checklist involving bias
assessment and quantitative synthesis are intended for meta-analyses so
did not apply to this study. Because we are not aware of any standard
procedures for conducting reviews and syntheses of methodology, no
protocol was prepared for this study.

@@

\textbf{Update numbers below, perhaps dynamically.}

The 25 included studies were analyzed based on 20 characteristics (see
Section \ref{sec:data}), and we report and describe some observations
here (Table \ref{tab:synthesis}). The studies were published in a
variety of journals, with some of greater frequency, though no single
journal published more than 3. The journal \emph{Evolving Systems}
published 2 of the included studies, \emph{Artificial Intelligence in
Medicine} published 3, and \emph{Hindawi - Journal of Healthcare
Engineering} published 3.

@@

We do observe one clear pattern: All of the proposed methods that most
evidently outperformed their comparators (Kohonen + CBR {[}20{]}, TWNFI
{[}18{]}, gravitational search algorithms (GSA) {[}26{]}, CBR + rules (+
DML) {[}28{]}, Gaussian process regression {[}30{]}, JITL-ELM {[}32{]})
are hybrids of localized modeling (in some cases CBR) with other
techniques, often metric learning. Though {[}27{]} and {[}15{]} report
non-superior performance by such hybrids, the pattern suggests the
importance of the similarity measure to the retrieval step. Meanwhile,
when proposed approaches targeted cohort demarcation or choice of
predictive model (statistical CBR {[}13{]}; individualized logistic
regression, decision tree, and random forest {[}14,16{]}), they did not
consistently outperform comparators.

@@

\textbf{Discuss public availability of code or implementations.}
\textbf{Methods, results, and inferential reproducibility
{[}\textbf{Matarese2022?}{]}.}

@@

@@

These programs used varying terminology for common concepts, and no
common term is in use for what we here refer to as localized modeling;
authors described their approaches as ``targeted prognosis'' {[}11{]},
``transductive inference'' {[}21{]}, ``personalized decision support''
{[}14{]}, ``personalized (predictive) models'' {[}26, in contrast to
``local models,''15,17,32{]}, and ``precision cohort'' analysis
{[}17,34,35{]}. We find uses of ``targeted'', ``personalized'', and
``precision'' generic and imprecise to this approach, while transductive
inference is an established term for a broader set of methods in ML.
Because \emph{local} naturally contrasts with \emph{global}, we propose
``localized models'' as a suitable term for this counterpart to
globally-fitted models.\footnote{We note, in response to one reviewer's
  comment, that for many if not most models used in ML, excepting GLMs,
  a predictor may play a different and more or less important role for
  some cases than others, as quantified by importance measures or
  model-agnostic explanations {[}39,40{]}. In this way, such models can
  also be viewed as ``localized''. We recognize that the use of global
  and local explanations is standard. As another reviewer pointed out,
  the terminology of local and global is also used in the unrelated
  context of federated ML to distinguish models trained using data
  stored in a single storage device (local) from their aggregations
  (global) {[}41,42{]}. The prospects for federated architecture to more
  effectively implement similarity cohort--based models are intriguing,
  and would require some reconciliation of terms. At present, we feel
  that these methods are sufficiently disjoint to avoid confusion in
  practice, but we suggest the more specific term ``similarity
  cohort--based models'' when the term ``local'' is overloaded.}

@@

The 25 studies included in our synthesis used composite techniques that
we organized into two major types. One type, similarity learning, was
used in 8 studies that tie in to the much larger literature on metric
learning {[}43{]}. These techniques are designed to detect structure in
data, especially associations between predictors and responses, and to
use this structure to inform the definition of a patient similarity
measure. Seven studies used response values in training data sets to
supervise similarity learning while two used unsupervised learning
(Table \ref{tab:composite}). Based on the key properties of metric
learning algorithms identified by {[}44{]}, most learned measures were
non-linear while some were local, and some were optimized globally while
others locally. However, not all qualified as metric learning, since the
resulting measures would not necessarily satisfy the triangle
inequality. Each then used its similarity measure to retrieve training
cases relevant to each testing case. Once retrieved, most took a
standard NN approach to generating predictions from the cohort; two
exceptions {[}29,34{]} proceeded to fit predictive models to the
retrieved cohorts and are discussed below.

@@

The similarity learning approaches took a variety of forms. Several
studies used unsupervised similarity learning, for example the
Mahalanobis distance {[}12{]} and the Kolmogorov entropy-based distance
{[}22{]}. In particular, {[}19{]} defined similarity as a weighted sum
of differences in predictor values and used linear regression to
optimize the weights for the predictive accuracy of the cohort they
retrieve. Others used supervised learning: {[}21{]} proposed an
iterative algorithm to optimize a set of fuzzy inference rules used to
retrieve relevant cases and generate a prediction, using
back-propagation on the rules' parameters. Their method was used in
several later studies returned by our search, which are not discussed
here because they did not originate the technique. {[}28{]} explicitly
appeal to a supervised distance metric learning (DML) technique {[}45{]}
to obtain a similarity measure on their set of dermoscopy and confocal
images that most effectively separates malignant from benign melanoma
tumors. We will discuss the remaining uses of supervised similarity
learning shortly.

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by {[}13{]} to optimize
the predictive performance of models fitted to cohorts retrieved using a
fixed similarity rather than count threshold, a counterpart to
conventional CBR they called ``statistical CBR''.\footnote{Drawing from
  {[}46{]}, we suggest the term ``combinatorial CBR'' for the
  then-conventional approach using similarity cohorts of fixed
  cardinality.} The step used a heuristic procedure to locate a
similarity threshold that (locally) maximizes predictive accuracy on the
training set. This is analogous to optimizing the neighborhood size
parameter in NN predictive modeling, so we will not discuss it further
except to mention that the authors found statistical CBR to outperform
conventional CBR on several data sets. {[}27{]} combined these learning
techniques: They used logistic regression on the training set to obtain
weights for the predictors (by predicting the binary outcome of kidney
transplant waitlist registration) and for the cases (by predicting
agreement of outcome between an index case and other training cases),
and they used exhaustion to optimize the size of the retrieved cohort
for the accuracy of the NN model using the previously optimized weights.

As noted above, these instances of similarity learning fit into a much
larger literature on metric learning, which has seen widespread use in
health informatics and modeling. That these studies satisfied our review
criteria is due to the novelty of the techniques at the time of
publication and the detailed attention paid to their techniques by the
authors. It also reflects a limitation of our study and of the
literature we set out to retrieve: We are aware of no standard terms in
use to identify the family of techniques that fit statistical models to
similarity-based cohorts. The retrieved studies whose use of such
techniques alone satisfied our criteria variably termed them
individualized, personalized, local, and patient-specific models, among
other terms (Table \ref{tab:composite}). We prefer the term localized
models, which is both sensitive and specific to uses of a similarity or
distance measure to retrieve a cohort to which a model is fit: It
includes some properly included techniques we did not have in mind
{[}29{]} yet excludes other techniques commonly referred to as
individualized, personalized, or precision.

@@

This relatively small sample exhibited a wide range of approaches, both
technically and conceptually. Discussion of the technical diversity of
these approaches has been summarized above and published in greater
detail in previous reviews {[}6,9,10{]}, and we focus here on the
conceptual. The design of a localized modeling approach can be
decomposed into three largely independent choices: (1) how to measure
the similarity between cases, (2) how to retrieve a cohort of cases
similar to an index case, and (3) how to generate a prediction or other
statistical insight for the index case from the similarity cohort. The
choices are the \emph{similarity measure}, the \emph{retrieval step},
and the \emph{statistical model}, respectively. These are diagrammed in
Figure \ref{fig:framework}.

@@

Each step can be unsupervised supervised: Unsupervised similarity
measures include discrete measures like the Levenshtein distance and
continuous measures like cosine similarity, while supervised measures
include the Mahalanobis distance and random forest proximity as well as
several composite measures with weights calculated from the data. Most
retrieval steps obtained cohorts with a uniform size (cardinality) or
similarity threshold, and the remaining were likewise only informed by
predictors; thus, while a retrieval step could in principle be
supervised, in this sample none were. Almost all models were supervised,
being that most performed predictive tasks, though {[}11{]} modeled only
survival rates within localized cohorts. Meanwhile, each step can be
optimized in a ML fashion by having its parameters tuned to improve
performance. We found no studies that tuned the calculation of
similarity, though most uniform cohort sizes were tuned. Some earlier
studies tuned generalized regression models, but more recent studies
have not.

@@

{[}13{]} & & Dermatology & Statistical CBR & 96.57\%\hspace{6em} & CBR &
95.43\%\hspace{6em} \\

@@

{[}18{]} & & Colon cancer diagnosis & TWNFI & 91.9\%\hspace{6em} & MLR
(Personalized) & 82.3\%\hspace{6em} \\

@@

{[}26{]} & & Colon cancer & knnGSA & 85.48\%\hspace{6em} & SVM &
82.14\%\hspace{6em} \\

@@

{[}14{]} & & 30-day mortality & Individualized LR & 0.830\hspace{6em} &
Individualized DC & 0.797\hspace{6em} \\

@@

{[}16{]} & & & Individualized LR & 0.824\hspace{6em} & Individualized DC
& 0.801\hspace{6em} \\

@@

{[}17{]} & & Diabetes & Personalized RF & 0.90\hspace{6em} & &

@@

{[}32{]} & & Discharge at 10 days & One-class JITL-ELM &

@@

\leavevmode\vadjust pre{\hypertarget{ref-Biecek2021}{}}%
\CSLLeftMargin{{[}39{]} }%
\CSLRightInline{Biecek P, Burzykowski T. Explanatory {Model Analysis}:
{Explore}, {Explain}, and {Examine Predictive Models}. {New York}:
{Chapman and Hall/CRC}; 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-Molnar2023}{}}%
\CSLLeftMargin{{[}40{]} }%
\CSLRightInline{Molnar C. Interpretable {Machine Learning}: {A Guide}
for {Making Black Box Models Explainable}. 2023.}

\leavevmode\vadjust pre{\hypertarget{ref-Moshawrab2023}{}}%
\CSLLeftMargin{{[}41{]} }%
\CSLRightInline{Moshawrab M, Adda M, Bouzouane A, Ibrahim H, Raad A.
Reviewing {Federated Machine Learning} and {Its Use} in {Diseases
Prediction}. Sensors 2023;23:2112.
\url{https://doi.org/10.3390/s23042112}.}

\leavevmode\vadjust pre{\hypertarget{ref-Brauneck2023}{}}%
\CSLLeftMargin{{[}42{]} }%
\CSLRightInline{Brauneck A, Schmalhorst L, Majdabadi MMK, Bakhtiari M,
Völker U, Baumbach J, et al. Federated {Machine Learning},
{Privacy-Enhancing Technologies}, and {Data Protection Laws} in {Medical
Research}: {Scoping Review}. Journal of Medical Internet Research
2023;25:e41588. \url{https://doi.org/10.2196/41588}.}

\leavevmode\vadjust pre{\hypertarget{ref-Yang2006}{}}%
\CSLLeftMargin{{[}43{]} }%
\CSLRightInline{Yang L. Distance {Metric Learning}: {A Comprehensive
Survey}. PhD thesis. Michigan State University, 2006.}

\leavevmode\vadjust pre{\hypertarget{ref-Bellet2014}{}}%
\CSLLeftMargin{{[}44{]} }%
\CSLRightInline{Bellet A, Habrard A, Sebban M. A {Survey} on {Metric
Learning} for {Feature Vectors} and {Structured Data} 2014.
\url{https://doi.org/10.48550/arXiv.1306.6709}.}

\leavevmode\vadjust pre{\hypertarget{ref-Xing2002}{}}%
\CSLLeftMargin{{[}45{]} }%
\CSLRightInline{Xing E, Jordan M, Russell SJ, Ng A. Distance {Metric
Learning} with {Application} to {Clustering} with {Side-Information}.
Advances in {Neural Information Processing Systems}, vol. 15, {MIT
Press}; 2002.}

\leavevmode\vadjust pre{\hypertarget{ref-Goyal2008}{}}%
\CSLLeftMargin{{[}46{]} }%
\CSLRightInline{Goyal N, Lifshits Y, Schütze H. Disorder inequality: A
combinatorial approach to nearest neighbor search. Proceedings of the
2008 {International Conference} on {Web Search} and {Data Mining}, {New
York, NY, USA}: {Association for Computing Machinery}; 2008, p. 25--32.
\url{https://doi.org/10.1145/1341531.1341538}.}

>>>>

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} (see Section \ref{sec:coherence}) that appears
to have been introduced independently several times. The approach fits
ML models, including regularized families of generalized regression
models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR
and NN prediction use patient similarity to obtain an ideally
homogeneous subpopulation matched to the index case, in localized
modeling one seeks a relevant yet heterogeneous subpopulation, each
member of which may deviate from the index case. A prediction or
inference is then obtained from a model fitted to this \emph{similarity
cohort}. The use of similarity cohorts distinguishes localized modeling
from classical GLMs that still dominate analysis in clinical trials and
comparative effectiveness research, as well as from many cutting-edge ML
models that are fitted to whole populations, while the cohorts'
heterogeneity and its utility to the model distinguish the approach from
CBR. In principle, this allows interpretable models like GLMs to
identify and measure the effects of determining factors that are
specific to the index case. In the reviewed studies, localized modeling
shows some promise toward simultaneously achieving the performance
improvements of ML and the interpretability of classical models.

@@

We include a PRISMA checklist as Supplemental Table 1 and a PRISMA
abstract checklist as Supplemental Table 2. Because we focus on
methodologies rather than conceptual approaches or evidence, we deviate
in some ways from PRISMA guidelines. In particular, those items of the
checklist involving bias assessment and quantitative synthesis are
intended for meta-analyses so did not apply to this study. Because we
are not aware of any standard procedures for conducting reviews and
syntheses of methodology, no protocol was prepared for this study.

@@

The 25 included studies were analyzed based on 20 characteristics (see
Section \ref{sec:data}), and we report and describe some observations
here (Table \ref{tab:synthesis}). The studies were published in a
variety of journals, with some of greater frequency, though no single
journal published more than 3. The journal \emph{Evolving Systems}
published 2 of the included studies, \emph{Artificial Intelligence in
Medicine} published 3, \emph{Hindawi - Journal of Healthcare
Engineering} published 3, and \emph{PLOS One} published 2.

@@

We do observe one clear pattern: All of the proposed methods that most
evidently outperformed their comparators (Kohonen + CBR {[}20{]}, TWNFI
{[}18{]}, gravitational search algorithms (GSA) {[}26{]}, CBR + rules (+
DML) {[}28{]}, Gaussian process regression {[}30{]}, JITL-ELM {[}32{]})
are hybrids of localized modeling (in some cases CBR) with other
techniques, often DML. Though {[}27{]} and {[}15{]} report non-superior
performance by such hybrids, the pattern suggests the importance of the
similarity measure to the retrieval step. Meanwhile, when proposed
approaches targeted cohort demarcation or choice of predictive model
(statistical CBR {[}13{]}; individualized logistic regression, decision
tree, and random forest {[}14,16{]}), they did not consistently
outperform comparators.

@@

\textbf{Discuss public availability of code or implementations.}
\textbf{Methods, results, and inferential reproducibility {[}39{]}.}

@@

\label{sec:coherence}

@@

These programs used varying terminology for common concepts, and no
common term is in use for what we here refer to as localized modeling;
authors described their approaches as ``targeted prognosis'' {[}11{]},
``transductive inference'' {[}21{]}, ``personalized decision support''
{[}14{]}, ``personalized (predictive) models'' {[}26, in contrast to
``local models,''15,17,32{]}, and ``precision cohort'' analysis
{[}17,34,35{]}. We find uses of ``targeted'', ``personalized'', and
``precision'' generic and imprecise to this approach, while transductive
inference is an established term for a broader set of methods in ML.
Because \emph{local} naturally contrasts with \emph{global}, we propose
``localized models'' as a suitable term for this counterpart to
globally-fitted models.\footnote{We note, in response to one reviewer's
  comment, that for many if not most models used in ML, excepting GLMs,
  a predictor may play a different and more or less important role for
  some cases than others, as quantified by importance measures or
  model-agnostic explanations {[}40,41{]}. In this way, such models can
  also be viewed as ``localized''. We recognize that the use of global
  and local explanations is standard. As another reviewer pointed out,
  the terminology of local and global is also used in the unrelated
  context of federated ML to distinguish models trained using data
  stored in a single storage device (local) from their aggregations
  (global) {[}42,43{]}. The prospects for federated architecture to more
  effectively implement similarity cohort--based models are intriguing,
  and would require some reconciliation of terms. At present, we feel
  that these methods are sufficiently disjoint to avoid confusion in
  practice, but we suggest the more specific term ``similarity
  cohort--based models'' when the term ``local'' is overloaded.}

@@

The 25 studies included in our synthesis used composite techniques that
we organized into two major types. One type, \emph{similarity learning},
was used in 8 studies that tie in to the much larger literature on DML
{[}44{]}. The other, \emph{cohort thresholding}, focused on choosing or
optimizing the manner in which cohorts were demarcated using the
similarity measure.

@@

The similarity learning approaches took a variety of forms. These
techniques are designed to detect structure in data, especially
associations between predictors and responses, and to use this structure
to inform the definition of a patient similarity measure. Seven studies
used response values in training data sets to supervise similarity
learning while two used unsupervised learning (Table
\ref{tab:composite}). Based on the key properties of DML algorithms
identified by {[}45{]}, most learned measures were non-linear while some
were locally learned, and some were optimized globally while others
locally. Not all qualified as DML, since the resulting measures would
not necessarily satisfy the triangle inequality. Each used its
similarity measure to retrieve training cases relevant to each testing
case. Once retrieved, most took a standard NN approach to generating
predictions; two exceptions {[}29,34{]} fit predictive models to the
retrieved cohorts.

Several studies used unsupervised similarity learning, for example the
Mahalanobis distance {[}12{]} and the Kolmogorov entropy-based distance
{[}22{]}. In particular, {[}19{]} defined similarity as a weighted sum
of differences in predictor values and used linear regression to
optimize the weights for the predictive accuracy of the cohort they
retrieve. Others used supervised learning: {[}21{]} proposed an
iterative algorithm to optimize a set of fuzzy inference rules used to
retrieve relevant cases and generate a prediction, using
back-propagation on the rules' parameters. Their method was used in
several later studies returned by our search, which are not discussed
here because they did not originate the technique. {[}28{]} explicitly
appeal to a supervised distance metric learning (DML) technique {[}46{]}
to obtain a similarity measure on their set of dermoscopy and confocal
images that most effectively separates malignant from benign melanoma
tumors. We will discuss the remaining uses of supervised similarity
learning shortly.

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by {[}13{]} to optimize
the predictive performance of models fitted to cohorts retrieved using a
fixed similarity rather than count threshold, a counterpart to
conventional CBR they called ``statistical CBR''.\footnote{Drawing from
  {[}47{]}, we suggest the term ``combinatorial CBR'' for the
  then-conventional approach using similarity cohorts of fixed
  cardinality.} The step used a heuristic procedure to locate a
similarity threshold that (locally) maximizes predictive accuracy on the
training set. This is analogous to optimizing the neighborhood size
parameter in NN predictive modeling, so we will not discuss it further
except to mention that the authors found statistical CBR to outperform
conventional CBR on several data sets. {[}27{]} combined these learning
techniques: They used logistic regression on the training set to obtain
weights for the predictors (by predicting the binary outcome of kidney
transplant waitlist registration) and for the cases (by predicting
agreement of outcome between an index case and other training cases),
and they used exhaustion to optimize the size of the retrieved cohort
for the accuracy of the NN model using the previously optimized weights.

As noted above, these instances of similarity learning fit into a much
larger literature on DML, which has seen widespread use in health
informatics and modeling. That these studies satisfied our review
criteria is due to the novelty of the techniques at the time of
publication and the detailed attention paid to their techniques by the
authors. It also reflects a limitation of our study and of the
literature we set out to retrieve: We are aware of no standard terms in
use to identify the family of techniques that fit statistical models to
similarity-based cohorts. The retrieved studies whose use of such
techniques alone satisfied our criteria variably termed them
individualized, personalized, local, and patient-specific models, among
other terms (Table \ref{tab:composite}). We prefer the term localized
models, which is both sensitive and specific to uses of a similarity or
distance measure to retrieve a cohort to which a model is fit: It
includes some properly included techniques we did not have in mind
{[}29{]} yet excludes other techniques commonly referred to as
individualized, personalized, or precision.

@@

This relatively small sample exhibited a wide range of approaches, both
technically and conceptually. Discussion of the technical diversity of
these approaches has been summarized above and published in greater
detail in previous reviews {[}6,9,10{]}, and we focus here on the
conceptual. The design of a localized modeling approach can be
decomposed into three largely independent choices: (1) how to measure
the similarity between cases, (2) how to retrieve a cohort of cases
similar to an index case, and (3) how to generate a prediction or other
statistical insight for the index case from the similarity cohort. The
choices are the \emph{similarity measure}, the \emph{cohort retrieval},
and the \emph{statistical model}, respectively. These are diagrammed in
Figure \ref{fig:framework}.

@@

Each step can be unsupervised supervised: Unsupervised similarity
measures include discrete measures like the Levenshtein distance and
continuous measures like cosine similarity, while supervised measures
include the Mahalanobis distance and random forest proximity as well as
several composite measures with weights calculated from the data. Most
retrieval steps obtained cohorts with a uniform size (cardinality) or
similarity threshold, and the remaining were likewise only informed by
predictors; thus, while a retrieval step could in principle be
supervised, in this sample none were. Almost all models were supervised,
being that most performed predictive tasks, though {[}11{]} modeled only
survival rates within localized cohorts. Meanwhile, each step can be
optimized in a ML fashion by having its parameters tuned to improve
performance. We found no studies that tuned the calculation of
similarity, though most uniform cohort sizes were tuned. Some earlier
studies tuned generalized regression models, but more recent studies did
not.

@@

{[}13{]} & Accuracy & Dermatology & Statistical CBR &
96.57\%\hspace{6em} & CBR & 95.43\%\hspace{6em} \\

@@

{[}18{]} & Accuracy & Colon cancer diagnosis & TWNFI &
91.9\%\hspace{6em} & MLR (Personalized) & 82.3\%\hspace{6em} \\

@@

{[}26{]} & Accuracy & Colon cancer & knnGSA & 85.48\%\hspace{6em} & SVM
& 82.14\%\hspace{6em} \\

@@

{[}14{]} & AUROC & 30-day mortality & Individualized LR &
0.830\hspace{6em} & Individualized DC & 0.797\hspace{6em} \\

@@

{[}16{]} & AUROC & & Individualized LR & 0.824\hspace{6em} &
Individualized DC & 0.801\hspace{6em} \\

@@

{[}17{]} & AUROC & Diabetes & Personalized RF & 0.90\hspace{6em} & &

@@

{[}32{]} & AUROC & Discharge at 10 days & One-class JITL-ELM &

@@

\leavevmode\vadjust pre{\hypertarget{ref-Matarese2022}{}}%
\CSLLeftMargin{{[}39{]} }%
\CSLRightInline{Matarese V. Kinds of {Replicability}: {Different Terms}
and {Different Functions}. Axiomathes 2022;32:647--70.
\url{https://doi.org/10.1007/s10516-021-09610-2}.}

\leavevmode\vadjust pre{\hypertarget{ref-Biecek2021}{}}%
\CSLLeftMargin{{[}40{]} }%
\CSLRightInline{Biecek P, Burzykowski T. Explanatory {Model Analysis}:
{Explore}, {Explain}, and {Examine Predictive Models}. {New York}:
{Chapman and Hall/CRC}; 2021.}

\leavevmode\vadjust pre{\hypertarget{ref-Molnar2023}{}}%
\CSLLeftMargin{{[}41{]} }%
\CSLRightInline{Molnar C. Interpretable {Machine Learning}: {A Guide}
for {Making Black Box Models Explainable}. 2023.}

\leavevmode\vadjust pre{\hypertarget{ref-Moshawrab2023}{}}%
\CSLLeftMargin{{[}42{]} }%
\CSLRightInline{Moshawrab M, Adda M, Bouzouane A, Ibrahim H, Raad A.
Reviewing {Federated Machine Learning} and {Its Use} in {Diseases
Prediction}. Sensors 2023;23:2112.
\url{https://doi.org/10.3390/s23042112}.}

\leavevmode\vadjust pre{\hypertarget{ref-Brauneck2023}{}}%
\CSLLeftMargin{{[}43{]} }%
\CSLRightInline{Brauneck A, Schmalhorst L, Majdabadi MMK, Bakhtiari M,
Völker U, Baumbach J, et al. Federated {Machine Learning},
{Privacy-Enhancing Technologies}, and {Data Protection Laws} in {Medical
Research}: {Scoping Review}. Journal of Medical Internet Research
2023;25:e41588. \url{https://doi.org/10.2196/41588}.}

\leavevmode\vadjust pre{\hypertarget{ref-Yang2006}{}}%
\CSLLeftMargin{{[}44{]} }%
\CSLRightInline{Yang L. Distance {Metric Learning}: {A Comprehensive
Survey}. PhD thesis. Michigan State University, 2006.}

\leavevmode\vadjust pre{\hypertarget{ref-Bellet2014}{}}%
\CSLLeftMargin{{[}45{]} }%
\CSLRightInline{Bellet A, Habrard A, Sebban M. A {Survey} on {Metric
Learning} for {Feature Vectors} and {Structured Data} 2014.
\url{https://doi.org/10.48550/arXiv.1306.6709}.}

\leavevmode\vadjust pre{\hypertarget{ref-Xing2002}{}}%
\CSLLeftMargin{{[}46{]} }%
\CSLRightInline{Xing E, Jordan M, Russell SJ, Ng A. Distance {Metric
Learning} with {Application} to {Clustering} with {Side-Information}.
Advances in {Neural Information Processing Systems}, vol. 15, {MIT
Press}; 2002.}

\leavevmode\vadjust pre{\hypertarget{ref-Goyal2008}{}}%
\CSLLeftMargin{{[}47{]} }%
\CSLRightInline{Goyal N, Lifshits Y, Schütze H. Disorder inequality: A
combinatorial approach to nearest neighbor search. Proceedings of the
2008 {International Conference} on {Web Search} and {Data Mining}, {New
York, NY, USA}: {Association for Computing Machinery}; 2008, p. 25--32.
\url{https://doi.org/10.1145/1341531.1341538}.}
